What We Eval in the Shadows

### 2021-08-10 paper/main.tex
changes to 4

@@

\begin{wrapfigure}{4}{5.8cm}  \small  \vspace*{-2mm}\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\#calls} & \#sites &
\multicolumn{3}{c}{\#calls} & \#sites \\\hline
\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
\end{tabular}

  \medskip  (a) All  \medskip  \medskip

  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{package_calls_per_run_per_call_site}

  (b) Small

\caption{Normalized calls} \label{cn}

\medskip\medskip
\begin{tabular}{c}
{\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
\end{tabular}
\caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
packages. Any run that does not call \eval is discarded, leaving us with
\packageNbruns runs. We recorded \packageAllcalls \eval calls originating from
\PkgHitEvalCallSites sites. This exercised \PkgHitEvalCallSitesAvgRatio of
sites, the coverage is similar to the package code coverage metric which is
\PkgCodeCoverage. The reason some sites are not exercised can be chalked down to
incomplete tests and analysis failures (\PkgFailedProgramsRatio of all scripts
crashed or timed out). Fig~\ref{fig:traced-eval-callsites} plots exercised site
with respect to all sites for each program; as can be seen coverage is unequal.
Figure~\ref{freq} summarizes the frequency of calls, number of calls on the left
and number of packages on the right. There are \packageFewcalls packages with
fewer than 100 calls, and \packageManycalls packages with more than 1,000 calls.
\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
over half of the observations. Fig.~\ref{tab:variantseval} summarizes use of
variants, the first row has sites (\emph{static}), the second has sites
encountered during analysis (\emph{exercised}) and the last gives calls
(\emph{invocations}). Most sites and calls go to \eval itself, \c{eval.parent}
is rare, and both \c{evalq} and \c{local} are barely used at all.

Figure~\ref{cn}(a) shows average count of calls from a given site and given run
and how many sites fall in that range. For instance, \packageRunbinh sites are
called over 2,000 times per run. Larger numbers suggest loops or recursive
contexts -- this seems to be the exception; \packageRunbina of sites are called
fewer than 50 times. Fig.~\ref{cn}(b) zooms in on low frequency sites. The
x-axis shows normalized calls, and the y-axis counts sites. Most sites are
invoked once, half as many are invoked twice, the drop off is sharp after that.

\Eval accepts any value as argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
representing a function calls), and \packageExpressionpercent are \c{expression}
objects (lists of expressions). Further inspection reveals that \c{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\c{ggplot2}.\footnote{To model inheritance in \c{ggproto}, one of the many
object-oriented systems for R.}

To estimate how much executable code is injected through \eval, we measure the
number of nodes in the expressions; for example, \c{x+1} counts as 3. We also
measured string lengths of unparsed expressions, but these measurements were
dominated by the size of data objects which could range in the MBs. The median
argument size is \packageMedianszeval (due to the symbols), and the average is
\packageAvgszeval nodes. The largest \eval input observed is
\packageMaxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
for legibility.

To estimate the work performed in \evals, we count instructions executed by the
interpreter. Most invocations perform relatively little work, with
\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
instructions; it is dominated by trivial symbol lookups. Figure~\ref{ev}(b) has
the work-intensive \evals, which go all the way to \packageMaxeventsRnd
instructions.

@@

\paragraph{Discussion.}
\Eval is used in CRAN packages. Nearly every fourth package contains at least
one \eval site. However, most packages that use \eval do that modestly. Half of
them include less than three \eval sites. Packages that rely heavily on \eval
are mostly the ones that provide statistical modeling and simulation functions.
At runtime, over a third of extracted programs triggered calls to \eval. Half
of the \packageAllcallsRnd recorded calls come from a single \eval site, but
for the rest, most of the packages have low \eval call frequency.

@@

\begin{figure}[h]
\begin{lstlisting}
.makeCppMethods <- function(methods, cppclassinfo, env) {
    cppMethods <- names(cppclassinfo@methods)
    newMethods <- names(methods)
    for(what in cppMethods[! cppMethods %in% newMethods])
        methods[[what]] <- eval(substitute(
                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
                                env)
    methods
}
\end{lstlisting}
  \caption{Rcpp::.makeCppMethods}
\end{figure}

@@

\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation make R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
this.

@@

text enclosed by c{\{} and c{\}} is
evaluated using \eval as an R expression and replaced by its result.

The ability to capture the current call as a language object (\code{match.call()}),
extract the unevaluated expression from function arguments (\code{substitute()})
and \eval with first class environment makes R the toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \code{data.table}
package defines a DSL for data frame manipulation using the subset operator
\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \code{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

Another example is the \code{tidyselect} package that defines a DSL for selecting
columns in a data frame powering number of data manipulation operation in the
\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
interpreter for evaluating column selectors such as \code{select(df,
starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
that starts with the string \code{length} and is not between \code{name} and
\code{mass} columns.

Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}. They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}). To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

>>>>

\medskip
  \medskip

@@

Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
packages. Any run that does not call \eval is discarded, leaving us with
\packageNbruns runs. In these runs, \packageAllcalls \eval calls were recorded
originating from \PkgHitEvalCallSites sites. The runs exercised
\PkgHitEvalCallSitesAvgRatio of sites, a ratio similar to the package code
coverage metric which is \PkgCodeCoverage. The reasons some sites are not
exercised can be chalked down to incomplete tests and analysis failures
(\PkgFailedProgramsRatio of the runs crashed or timed out).
Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
sites for each program; as can be seen coverage is unequal.

Fig.~\ref{freq} summarizes the frequency of calls, call counts on the left and
number of packages on the right. There are \packageFewcalls packages with fewer
than 100 calls, and \packageManycalls packages with more than 1,000 calls.
\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
over half of our observations.

Fig.~\ref{tab:variantseval} summarizes use of variants, the first row has sites
(\emph{static}), the second has sites encountered during analysis
(\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
calls go to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
\c{local} are barely used at all.

\begin{wrapfigure}{r}{5cm}  \small  \vspace*{-2mm}\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\#calls} & \#sites &
\multicolumn{3}{c}{\#calls} & \#sites \\\hline
\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
\end{tabular}
\caption{Normalized calls} \label{cn}
\end{wrapfigure}

Fig.~\ref{cn} shows average count of calls from a given site and given run and
how many sites fall in that range. For instance, \packageRunbinh sites are
called over 2,000 times per run. Larger call counts suggest the presence of
loops or recursive functions, but given the data this seems to be the exception;
\packageRunbina of sites are called fewer than 50 times, most sites are invoked
once and half as many are invoked twice.

\Eval accepts any value as argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
representing a function calls), and \packageExpressionpercent are \c{expression}
objects (lists of expressions). Further inspection reveals that \c{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\c{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
\c{ggproto}, one of the many object-oriented systems for R.}

\begin{wrapfigure}{r}{5.8cm}
\begin{tabular}{c}
{\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
\end{tabular}
\caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

To estimate how much executable code is injected through \eval, we measure the
size of the arguments in terms of number of nodes. For example,
\c{expression(x+1)} has a size of 3. The large number of symbols skew the median
to \packageMedianszeval but the average size is \packageAvgszeval nodes. The
largest \eval input we observed was an expression of \packageMaxszeval nodes, a
significant chunk of code. Fig.~\ref{fig:sizedistribution} shows the
distribution of sizes for arguments of fewer than 25 nodes. Sizes drop rapidly,
there are few observations larger than 15 nodes. The long tail is omitted for
legibility. As an alternative to counting nodes, we tried measuring string
lengths of expressions after calling a function to deparse values back to
strings, but these measurements were dominated by massive data objects which
could range in the MBs.

To assess how much computational work is performed in \evals, we count the
instructions executed by the interpreter. Most call do relatively little, with
\packageSmalleventspct of calls executing fewer than 50 instructions. The violin
plot of Fig.~\ref{ev}(a) shows the distribution of short running \evals; the
data is dominated by \evals which simply perform symbol lookup. Fig.~\ref{ev}(b)
shows the distribution of work-intensive \evals which go all the way to
\packageMaxeventsRnd instructions.

@@

\paragraph{Discussion.}
\Eval is widely used in CRAN packages. Nearly every fourth package contains at
least one \eval call site. However, most packages use \eval modestly. Packages
that rely heavily on \eval are mostly ones that provide statistical modeling and
simulation functions. At runtime, over a third of our scripts triggered calls to
\eval, note that all scripts performed \eval originating from base package but
these are disregarded here. Half of the calls come from a single site, for the
rest, most packages have low call frequency. The data is consistent with usage
of \eval for configuration and meta-programming and not in performance sensitive
contexts such as loops. While most arguments to \eval are small, often just a
variable name, and do little work, we have also observed large \eval arguments
and arguments that perform massive amounts of work.

@@

\begin{figure}[h]
\begin{lstlisting}
.makeCppMethods <- function(methods, cppclassinfo, env) {
    ...
    for(what in cppMethods[! cppMethods %in% newMethods])
        methods[[what]] <- eval(substitute(
                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))), env)
    ...
}
\end{lstlisting}
  \caption{Rcpp::.makeCppMethods}
\end{figure}

@@

\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation makes R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
\code{glue} function in action.

@@

The \code{glue} function extracts snippets of R code enclosed between \c{\{} and
\c{\}}, evaluates them using \eval, and splices the result to construct the
final string.

Another example is the \code{data.table} package that defines a DSL for data
frame manipulation using the subset operator \code{[}. This subset operation is
defined as \code{DT[i, j, k]}, where expression \code{j} is used to calculate on
a subset/reorder of rows from \code{i} grouped by \code{k}. This enables compact
queries on data frames such as \code{flights[carrier == "AA", .N, by = origin]}
which will compute the number of flights done by American Airlines (\code{"AA"})
for each origin airport. At its core, \code{[} uses \code{substitute} to capture
the operand expressions, transforms them, and evaluates them using \eval in an
environment which has data frame columns available as bindings.
